Litrepl: Literate Paper Processor Promoting Transparency More Than Reproducibility

### 2025-01-02 paper/paper.tex
Update paper

\documentclass[a4paper,12pt,twocolumn]{article}

@@

The concept of Literate Programming, initially proposed by Donald Knuth,
revolves around explaining to human beings what we want a computer to do. This
approach is embodied in tools like WEB, CWEB, and NOWEB, which focus on
non-interactive, ahead-of-time compilation. The process includes two primary
modes of operation: weaving, which creates human-readable documentation, and
tangling, which generates machine-executable code. Over time, the concept has
evolved, showing a trend towards simplification \cite{Knuth1984lp,
Ramsey1994lps}.

Concurrently, a concept of human-computer interaction called the
Read-Evaluate-Print Loop or ''REPL'' gained traction, notably within the LISP
an APL communities.\cite{Spence1975apl, McCarthy1959recfun, Iverson1962apl}

@@

A significant milestone in this area became the IPython terminal, which later
evolved into the Jupyter Project. The creators introduced a new document format
called the Notebook, characterized by a series of logical sections of various
types, including text and code, which could directly interact with programming
language interpreters. This interactive communication, akin to REPL style
programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
Literate Computing, which includes goals of spanning a wide audience range,
boosting reproducibility, and fostering collaboration. To achieve these
objectives, several technical decisions were made, notably the introduction of
bidirectional communication—between the computational core, known as the Jupyter
Kernel, and the Notebook acting as a client, along with another layer of
client-server communication between the Notebook web server and the user’s web
browser \cite{Perez2007IPython, Granger2021litcomp, Kluyver2016jupnb}.

While we recognize the importance of these goals within the Literate Computing
framework, we argue that reproducibility is paramount. Successfully addressing
reproducibility would alone suffice to enhance communication over time among
research project participants and significantly expand the audience.  However,
as it became clear (\cite{Dolstra2010, Vallet2022}), this challenge extends far
beyond the scope of a human-computer interaction system, and even beyond the
typical boundaries of software distribution management for a particular
programming language. A comprehensive solution to the software deployment
problem should operate at the entire operating system level.

@@

We accomplish this through the following steps. First, we utilize simple
bidirectional text streams for inter-process communication with language
interpreters to evaluate code. Second, we advocate for the reuse of existing
text document formats. In both the Markdown and \Latex evaluators that we have
implemented, simplified parsers are utilized to distinguish code and result
sections from the rest of the document. As of now, we support Python and Shell
interpreter families, as well as a custom AI communication interpreter. Finally,
we strive to leverage POSIX \cite{POSIX2024} operating system facilities as much
as possible.

@@

Template grammars similar to the illustrative example in Figure
\ref{fig:document-representation} are encoded for Markdown and \Latex formats.
Before each run, Litrepl calls Lark\cite{Lark} to compile a customized parser
and uses it to access the sections.

>>>>

\documentclass[letterpaper,12pt,twocolumn]{article}

@@

The concept of Literate Programming, initially proposed by Donald Knuth,
revolves around explaining to human beings what we want a computer to do. This
approach is embodied in tools like WEB, CWEB, and NOWEB, which focus on
non-interactive, ahead-of-time compilation. The process includes two primary
modes of operation: weaving, which creates human-readable documentation, and
tangling, which generates machine-executable code. Over time, the concept has
evolved, showing a trend towards simplification \textcite{Knuth1984lp},
\textcite{Ramsey1994lps}.

Concurrently, a concept of human-computer interaction called the
Read-Evaluate-Print Loop or ''REPL'' gained traction, notably within the LISP an
APL communities.\textcite{Spence1975apl, McCarthy1959recfun, Iverson1962apl}

@@

A significant milestone in this area became the IPython terminal, which later
evolved into the Jupyter Project. The creators introduced a new document format
called the Notebook, characterized by a series of logical sections of various
types, including text and code, which could directly interact with programming
language interpreters. This interactive communication, akin to REPL style
programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
Literate Computing, which includes goals of spanning a wide audience range,
boosting reproducibility, and fostering collaboration. To achieve these
objectives, several technical decisions were made, notably the introduction of
bidirectional communication—between the computational core, known as the Jupyter
Kernel, and the Notebook acting as a client, along with another layer of
client-server communication between the Notebook web server and the user’s web
browser \textcite{Perez2007IPython, Granger2021litcomp, Kluyver2016jupnb}.

While we recognize the importance of these goals within the Literate Computing
framework, we argue that reproducibility is paramount. Successfully addressing
reproducibility would alone suffice to enhance communication over time among
research project participants and significantly expand the audience.  However,
as it became clear (\textcite{Dolstra2010, Vallet2022}), this challenge extends
far beyond the scope of a human-computer interaction system, and even beyond the
typical boundaries of software distribution management for a particular
programming language. A comprehensive solution to the software deployment
problem should operate at the entire operating system level.

@@

We accomplish this through the following steps. First, we utilize simple
bidirectional text streams for inter-process communication with language
interpreters to evaluate code. Second, we advocate for the reuse of existing
text document formats. In both the Markdown and \Latex evaluators that we have
implemented, simplified parsers are utilized to distinguish code and result
sections from the rest of the document. As of now, we support Python and Shell
interpreter families, as well as a custom AI communication interpreter. Finally,
we strive to leverage POSIX \textcite{POSIX2024} operating system facilities as
much as possible.

@@

Template grammars similar to the illustrative example in Figure
\ref{fig:document-representation} are encoded for Markdown and \Latex formats.
Before each run, Litrepl calls Lark\textcite{Lark} to compile a customized
parser and uses it to access the sections.
